The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2010-12-02 smurf_paper.tex
flesh-out text describing S2SRO data including time series and PSDs

\begin{abstract}
  We describe the properties of data from the Submillimetre Common
  User Bolometer Array 2 (SCUBA-2), and the production of maps using
  the Sub-Millimetre User Reduction Facility (SMURF).
\end{abstract}

@@

The load on the detectors can vary substatially, both from thermal
sources within and outside the SCUBA-2 cryostat. In order to operate
the camera across a broad dynamic range of conditions, it is therefore
necessary to thermally control the focal plane. This is achieved by
providing a fridge {\em base temperature}, $T_\mathrm{B}$ that is
well-below $T_\mathrm{C}$, and then using heaters that are connected
to each bolometer to place the array in to transition.

@@

For further detail see...

\subsubsection{Scan strategy}

@@

This paper will be focussed on S2SRO: only one subarray at each
wavelength; fridge oscillations; working detectors (and variation from
setup-to-setup).

Show some time series.

Show power spectra.

Show cross-spectra to emphasize how correlated things are?

@@

@@

\subsubsection{Fridge oscillations}
As shown in Fig.~\ref{fig:bolos_mix}, the dominant signal in both
bands is clearly an oscillation in the fridge base temperature as
evidenced by the low-resolution mixing chamber thermometer. The total
power delivered to the bolometers is comparable at each wavelength,
further suggesting a source internal to the cryostat, rather than an
external optical load. Finally, there is no obvious correspondance
with either the azimuth or elevation motion.

Show what's left over when we remove the common-mode, and the
correlation of the resulting signal with patterns in $G$ and
$T_\mathrm{C}$ across the array.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pca.pdf}
\caption{The first six modes from a principal component analysis,
  ordered by decreasing significance, of the combined 450 and
  850\,\micron\ time-series bolometer data. For each component, the
  top plot shows the time series of the eigenvector, normalized by its
  \rms. The bottom coloured panels show the eigenvalues for the
  bolometers at each wavelength (the amplitude of the eigenvector in
  the time series). For reference, both the mean, $\bar{\lambda}$, and
  \rms, $\lambda_\mathrm{rms}$, eigenvalues for the bolometers in each
  subarray are also shown. The first two components have comparable
  amplitudes, and dominate by at least a factor of 10 any other
  components of the time series. We suspect that they are caused by
  atmospheric variations (expected to be brighter at 450\,\micron),
  and fridge oscillations (expected to be comparable at each
  wavelength). However, the two eigenvectors do not necessarily map
  directly to these physical sources under this interpretation; rather
  they are two \emph{different} linear combinations of the underlying
  physical signals that give rise to statistically uncorrelated
  components. The third component accurately reproduces an apparent
  data-acquisition glitch that appears only at 450\,\micron\ (see
  Section ???). The fourth component is a ...\,Hz line feature only
  observed in column 15 of the 450\,\micron\ subarray. The fifth
  component shows a series of steps and glitches that appear
  predominantly along the bottom-left edge of the 850\,\micron\
  subarray where the bolometers are only tenuously locked. Finally,
  the sixth component is degenerating into a less significant, and
  more randomly scattered signal across the focal plane. Subsequent
  components follow this trend, and would generally be kept in a
  PCA-cleaning procedure.}

\label{fig:pca}
\end{figure*}

>>>>

\begin{abstract}
  We describe the properties of data from the Submillimetre Common
  User Bolometer Array 2 (SCUBA-2) taken during the Shared Risk
  Observing period from 2010 February 11 to 2010 March 25, and the
  production of maps using the Sub-Millimetre User Reduction Facility
  (SMURF). Our iterative approach is similar to those used to reduce
  data from other ground-based submillimetre telescopes in recent
  years, and recovers angular scales up to the size of the array
  footprint (approximately 2\,arcmin). The method is also able to
  achieve the theoretical white-noise limit of the instrument for
  point-source studies.
\end{abstract}

@@

The load on the detectors can vary substantially, both from thermal
sources within and outside the SCUBA-2 cryostat. In order to operate
the camera across a broad dynamic range of conditions, it is therefore
necessary to thermally control the focal plane. This is achieved by
providing a fridge {\em base temperature}, $T_\mathrm{B}$ that is
well-below $T_\mathrm{C}$, and then using heaters that are connected
to each bolometer to place the array in to transition.

@@

In general, bolometers have two predominant noise components: (i) the
effectively white thermal noise produced by the phonon and optical
loading on the absorbers; and (ii) and longer-timescale drift
resulting from slow variations in the mean load (e.g. fridge,
telescope, or atmospheric changes), and also the readout
electronics. As such, the spectrum corresponding to a bolometer
time-series exhibits a ``$1/f$'' shape below some knee frequency,
$f_\mathrm{knee}$, and is relatively flat at higher frequencies. Other
smaller effects may include line features in the spectrum (a common
culprit is pickup from AC mains, or microphonic signals induced by
mechanical vibrations), and finally a roll-off towards the Nyquist
frequency caused by the anti-aliasing filter that precedes
digitization.

In order to observe astronomical sources, it is clearly advantageous
that the signals appear in the region of the spectrum that offers the
highest \snr, namely frequencies above $f_\mathrm{knee}$. This effect
is achieved by modulating the signal; the original SCUBA primarily
used the JCMT chopping secondary to place sources into the spectrum at
about $\sim$10\,Hz. The problem with this strategy was that SCUBA was
ultimately limited to recovering information on angular scales smaller
than the chop throw (a maximum of a couple of arcmin)... cite
somebody. The primary approach to map-making with SCUBA-2 is instead
to slew the entire telescope at such a rate that angular scales of
interest are crossed on time scales shorter than $1/f_\mathrm{knee}$,
similar to other ground-based instruments (MAMBO, BOLOCAM, SHARC-II,
AzTEC, LABOCA,...). The desire to scan rapidly has two basic
constraints: (i) the telescope cannot physically move faster that
several hundred arcsec\,sec$^{-1}$ without requiring unacceptably
large accelerations to change direction, and (ii) bolometers are
designed with a thermal time constant that balance response with
sensitivity. Taking these constraints into consideration, SCUBA-2 was
ultimately designed to work at a maximum scan velocity of
600\,arcsec\,sec$^{-1}$. The data acquisition system has a target
sample rate of 200\,Hz, which at that scan velocity results in a
sample every 3.0\,arcsec, or approximately one third of the
450\,\micron\ diffraction-limited full-width at half-maximum (FWHM) --
a typical rule-of-thumb for adequately sampling a Gaussian point
spread function (PSF). Similarly, the SCUBA-2 bolometers are designed
to have a time constant of 2 ms \textbf{check that number!} in order
to avoid distorting PSF-sized sources at this maximum scan rate.

The scan strategy is designed, primarily, to mitigate the effects of
$1/f$ drift. The simplest scan one might use is a raster pattern, or a
series of parallel passes across this field. However, such a scan
yields extremely poor performace for structures that are transverse to
the scan direction, since a long time will have ellapsed between each
parallel step, and the signal will therefore be totally dominated by
the $1/f$ noise. In addition, other noise features may appear in the
data periodically. With such a regularly spaced periodic scan
strategy, these noise features can easily conspire with the scan
pattern to produce similarly periodic noise structures in the map, or
\emph{scan-synchronous} noise. It has therefore been recognized for
some time that scan strategies that visit each region of the map on as
many different time scales, and scan angles as possible offer the best
protection against large-scale noise (...cite some people...). As a
tradeoff between truly random scan patterns and feasibility of
implementation taking into consideration both mechanical and
scheduling constraints, SCUBA-2 presently has three primary scan
modes: PONG (cite Borys et al.), Lissajous, and Daisy... (more
citations? Kovacs et al. has a nice description... also CMB
literature).

@@

The S2SRO period, as mentioned above, took place during February and
March 2010. For the purpose of this paper, we will consider data taken
as early as 2009 December 3, at which point the instrument was being
commissioned in a similar configuration to that of the S2SRO
period. At this time, each of the 450 and 850\,\micron\ focal planes
were populated with single subarrays, s4a and s8d respectively. There
were two main mechanical problems to contend with during this
period. First, the fridge was not able to cool the focal plane to the
desired base temperature of ... which affected stability, particularly
at 850\,\micron. Second, there were problems with the subarray
fabrication process which led to large numbers of dead pixels, as well
as large-scale variations in $T_\mathrm{C}$ and the thermal
conductivity, $G$. These latter issues made it difficult to
simulataneously lock all of the bolometers, and in addition many of
the bolometers that were locked did not achieve their minimum
theoretical noise performance, since it is not possible to adjust the
individual temperature biases and place each device into the optimal
region of the superconducting transition.

While there were moderate variations from night-to-night during the
observing period, there were typically 600 to 900 working detectors in
each of the s4a and s8d subarrays. Fig.~\ref{fig:sensitivities} shows
representative responsivity maps for the two subarrays from the middle
of S2SRO based on the response to ramping the bolometer heaters.

\begin{figure}
\vspace{2cm}
\textbf{This will be a figure showing s4a and s8d responsivity maps}
\vspace{2cm}
\caption{The response of each bolometer to known input heater power
  for a typical night during S2SRO. The s4a subarray had a
  particularly large variation in $T_\mathrm{C}$ towards the
  bottom-right which resulted in lower responsivity, and
  correspondingly larger noise. The s8d array also suffered
  significant variations across the focal plane, in particular
  removing larger number of useful detectors from the bottom-left.}
\label{fig:sensitivities}
\end{figure}

\subsubsection{Time series}

In Fig.~\ref{fig:bolos_mix} we show sample time series from single
bolometers in each of the subarrays, as well as variations in the
mixing chamber temperature (as a proxy for the focal-plane
temperature), and the telescope pointing during a PONG scan.

@@

Both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, from $-0.04$\,pW to $+0.08$, are larger
than the $-0.03$\,pW to $+0.05$\,pW fluctuations at 850\,\micron. Such
behaviour might be expected if there is a comparable varying thermal
load from the fridge at each wavelength, but a larger contribution of
astmospheric variations through the 450\,\micron\ bandpass filters. We
also note that there is no obvious strong correlation between the
low-frequency signal structure at either wavelength at the telescope
motion.

The low-frequency signal component of the bolometer signals is also
highly correlated amongst bolometers in the same subarray. We have
calculate a common-mode signal, $\mathbf{c}(t)$ as the straight
average time series of all the working bolometers. We then fit the
amplitude of $\mathbf{c}(t)$ at each wavelength to the time series
shown in Fig.~\ref{fig:bolos_mix} and remove it, yielding the grey
residual signals. These residuals are quite flat, although still with
noticeable variations. The white noise is also apparent, and larger at
450\,\micron\ as one would expect from the larger loading compared to
850\,\micron.

\subsubsection{Power spectral densities}

Next, we produce power spectral density (PSD) plots for several
representative bolometers in Fig.~\ref{fig:pspec}. To produce this
figure, we follow the convention that the PSD as a function of
frequency, $\mathbf{P}(f)$, is normalized such that the integral over
frequency gives the same variance as the time-series variance across
the full time series. In other words, given a bolometer signal
$\mathbf{b}(t)$,
%
\begin{equation}
\int \mathbf{P}(f) df = \langle\mathbf{b}^2(t)\rangle.
\end{equation}
%
However, we only show the PSD up to the Nyquist frequency, so the
conversion from variance measured in the frequency domain to the time
domain requires and additional multiplcation by 2; hence the units are
pW$^2$\,Hz$^{-1}$ rather than pW$^2$\,s.

@@

The dotted coloured lines show the PSDs for raw, though flatfielded
data. At each wavelength, there is a clear $1/f$ knee at a few Hz,
followed by a predominantly white spectrum punctuated by line features
above $\gsim 30$\,Hz, and finally roll-off caused by the anti-aliasing
filter above $\gsim 70$\,Hz. As indicated in the previous section, the
correlation between the low-frequency components of the different
bolometer signals is large. The solid black lines in
Fig.~\ref{fig:pspec} indicate the PSDs of the common-modes
$\mathbf{c}(t)$ at each wavelength, which reproduce most of the
low-frequency structures, as well as the high-frequency line
features. The $\mathbf{c}(t)$ otherwise drop substantially below the
individual bolometer PSDs at high-frequency as expected if the
bolometers are dominated by un-correlated white noise. Finally, we
note that the amplitudes and slopes of the $\mathbf{c}(t)$ are
similar at both 450 and 850\,\micron. However, since the
450\,\micron\ white noise is larger than at 850\,\micron\ (shown
approximately by the horizontal dotted lines), the $1/f$ knee occurs
at a \emph{lower} frequency at 450\,\micron.

Next, the common-mode signals are fit to each bolometer time series
and removed as in Fig.~\ref{fig:bolos_mix}, and the resulting PSDs are
shown as solid coloured lines. In this example, the residual signals
have $1/f$ knees significantly lower than in the raw PSDs.

For reference, the top horizontal axis has been converted to angular
scale assuming a scan velocity of 200\,arcsec\,sec$^{-1}$ which was
typical of the S2SRO observations. The power spectra of the PSFs in
each band (arbitrary normalized to peak values of $10^{-6}$ and
$10^{-7}$\,pW$^2$\,Hz$^{-1}$ at 450 and 850\,\micron\ respectively)
are also showed as dashed lines for this assumed scan velocity showing
that the smallest features resolvable by the telescope may be slightly
affected by the excess noise in the line features. At the lower
frequency end, the $1/f$ noise dominates at scales $\gsim 2$\,arcmin
in the raw data and $\gsim 10$\,arcmin in the common-mode cleaned data
at 450\,\micron, and at scales $\gsim 1$\,arscmin in the raw data and
$\gsim 2$\,arcmin in the common-mode cleaned data at 850\,\micron.

\subsubsection{Principal component analysis}

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pca.pdf}
\caption{The first six modes from a principal component analysis of
  the combined 450 and 850\,\micron\ time-series bolometer data.}
\label{fig:pca}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pca.pdf}
\caption{The first six modes from a principal component analysis,
  ordered by decreasing significance, of the combined 450 and
  850\,\micron\ time-series bolometer data. For each component, the
  top plot shows the time series of the eigenvector, normalized by its
  \rms. The bottom coloured panels show the eigenvalues for the
  bolometers at each wavelength (the amplitude of the eigenvector in
  the time series). For reference, both the mean, $\bar{\lambda}$, and
  \rms, $\lambda_\mathrm{rms}$, eigenvalues for the bolometers in each
  subarray are also shown. The first two components have comparable
  amplitudes, and dominate by at least a factor of 10 any other
  components of the time series. We suspect that they are caused by
  atmospheric variations (expected to be brighter at 450\,\micron),
  and fridge oscillations (expected to be comparable at each
  wavelength). However, the two eigenvectors do not necessarily map
  directly to these physical sources under this interpretation; rather
  they are two \emph{different} linear combinations of the underlying
  physical signals that give rise to statistically uncorrelated
  components. The third component accurately reproduces an apparent
  data-acquisition glitch that appears only at 450\,\micron\ (see
  Section ???). The fourth component is a ...\,Hz line feature only
  observed in column 15 of the 450\,\micron\ subarray. The fifth
  component shows a series of steps and glitches that appear
  predominantly along the bottom-left edge of the 850\,\micron\
  subarray where the bolometers are only tenuously locked. Finally,
  the sixth component is degenerating into a less significant, and
  more randomly scattered signal across the focal plane. Subsequent
  components follow this trend, and would generally be kept in a
  PCA-cleaning procedure.}
\label{fig:pca}
\end{figure*}
